---
title: "Project Appendix"
subtitle: "STAT 511"
author: "Moses Sanches, Shuyu Wang, Theodore Choi"
date: "`r Sys.Date()`"
format: 
  html:
    self-contained: true
    toc: true
    toc_float: true
    number-section: false
    highlight: tango
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE)

```

```{r}
#| message: false
#| warning: false
# Libraries we used
library(openintro)
library(tidyverse)
library(dplyr)
library(scales)
library(leaps)
library(MASS)
library(olsrr)
library(car)
library(broom)
library(pastecs)
library(countrycode)
```

### Preliminary analysis using basic visualizations

```{r}
# load data (must have file in correct project folder)
life_exp <- read.csv("./data/Life Expectancy Data.csv")

# add continent to dataset
# https://stackoverflow.com/questions/47510141/get-continent-name-from-country-name-in-r
life_exp$Continent <- countrycode(sourcevar = life_exp[,"Country"],
                                  origin = "country.name",
                                  destination = "continent")

# set response variable
y <- life_exp$Life.expectancy

```

```{r}
# getting summaries of the status counts
# string values did not return summary statistics
summary_num <- stat.desc(select_if(life_exp, is.numeric))
round(summary_num, 2)

# frequency of status by country
table(life_exp$Continent, life_exp$Status) 
```

### Producing visual summaries

```{r}
# Getting counts of the Status (Developing vs developed)
counts_status <- table(life_exp$Status)

# Plotting a bar chart of the counts
barplot(counts_status,
        main = "Status Counts",
        ylab = "Counts",
        xlab = "Status")

# Getting frequency of Adult Mortality
# Defined as:
# "Adult Mortality Rates of both sexes
# (probability of dying between 15 and 60 years per 1000 population)"
hist(life_exp$Adult.Mortality,
     main = "Histogram of Adult Mortality",
     xlab = "Adult Mortality")

# Consumption of Alcohol per capita (15+) in liters
hist(life_exp$Alcohol,
     main = "Histogram of Alcohol",
     xlab = "Alcohol Consumption")

# Hepatitis B (HepB) immunization coverage among 1-year-olds (%)
hist(life_exp$Hepatitis.B,
     main = "Histogram of Hepatits B",
     xlab = "Infected with Hepatitis B")

# Year
hist(life_exp$Year,
     main = "Histogram of year",
     xlab = "Years")

# Plotting a bar chart of the  Continent counts
# (too many countries to plot cleanly)
counts_continent <- table(life_exp$Continent)
barplot(counts_continent,
        main = "Count of Continents",
        ylab = "Counts",
        xlab = "Continents")

# infant deaths
hist(life_exp$infant.deaths,
     main = "Histogram of Infant Deaths",
     xlab = "Infant Deaths")

# percentage expenditure 
hist(life_exp$percentage.expenditure,
     main = "Histogram of Percentage Expenditure",
     xlab = "Percent Expenditure")

# Measles
hist(life_exp$Measles,
     main = "Histogram of Measles",
     xlab = "Infected with Measles")

# BMI
hist(life_exp$BMI,
     main = "Histogram of BMI",
     xlab = "BMI")

# under 5 deaths
hist(life_exp$under.five.deaths,
     main = "Histogram of Under 5 Deaths",
     xlab = "Amount of deaths under 5")

# Polio
hist(life_exp$Polio,
     main = "Histogram of Polio Occurence",
     xlab = "Polio Occurrence")

# Total Expenditure
hist(life_exp$Total.expenditure,
     main = "Histogram of Total Expenditure",
     xlab = "Total Expenditure")

# Diphtheria
hist(life_exp$Diphtheria,
     main = "Histogram of Frequency of Diphtheria",
     xlab = "Diphtheria")

# AIDS
hist(life_exp$HIV.AIDS,
     main = "Histogram of Frequency of HIV AIDS",
     xlab = "HIV AIDS")

# GDP
hist(life_exp$GDP,
     main = "Histogram of GDP",
     xlab = "GDP")

# Population
hist(life_exp$Population,
     main = "Histogram of Population",
     xlab = "Population")

# Thinness(1-19 years):
# Prevalence of thinness among children and adolescents for Age 10 to 19 (%)
hist(life_exp$thinness..1.19.years,
     main = "Histogram of Thinness from age 10 to 19",
     xlab = "Age")

# Thinness(5-9 years):Prevalence of thinness among children for Age 5 to 9(%)
hist(life_exp$thinness.5.9.years,
     main = "Histogram of Thinness from age 5 to 9",
     xlab = "Age")

# Income Composition of Resources:
# Human Development Index in terms of income composition of resources
# (index ranging from 0 to 1)
hist(life_exp$Income.composition.of.resources,
     main = "Histogram of Resource Composition",
     xlab = "Index")

# Schooling:Number of years of Schooling(years)
hist(life_exp$thinness.5.9.years,
     main = "Histogram of Years of Schooling",
     xlab = "Years")

# plotting a histogram of the response variable (life expectancy)
hist(y,
     main = "Histogram of Life Expectancy",
     xlab = "Age")

# boxplot of life expectancy across the continents
life_exp %>%
  ggplot(aes(x = Continent,
             y = Life.expectancy,
             fill = Continent)) +
  geom_boxplot(outlier.color = "black",
               outlier.shape = 16,
               outlier.size = 2,
               notch = FALSE)
```

> The data still has some missing values and can be a little difficult to interpret - use of the data descriptors found in the link is highly recommended. The Life Expectancy column (response variable) is skewed left, so some transformations may have to be performed to ensure a normal curve. We can also see skewed distributions across the variables (Histogram of Infant Deaths and Histogram of Under 5 Deaths among others), possible missing data (as seen in Histogram of Total Expenditure), as well as possible outliers (as seen in Histogram of Population,the Histograms of thinness). Another interesting issue that we can see is a bimodal distribution (as seen in the Histogram of BMI). There may also be issues with the frequency of the number of years for the early 2000s as there is a lot of data for 2000.

### Getting best models

```{r}

# import data
filepath_2013 <- "./data/Longevity Data Year 2013 Cleaned Transformed.rds"

# read in the cleaned and transformed data
life_exp_trnsfrm <- readRDS(filepath_2013)

```

```{r}
#| echo: false
# Stepwise Backwards/Forward/Both Regression

# using stepwise functions to determine the best models
Base <- lm(longevity ~ 1, data = na.omit(life_exp_trnsfrm))
FullModel <- lm(longevity ~ ., data = na.omit(life_exp_trnsfrm))

# using forward progression
step(Base, scope = list(upper = FullModel, lower = Base),
     direction = "forward", trace = FALSE)

# using backwards progression
step(FullModel, direction = "backward", trace = FALSE)

# using both
step(Base, scope = list(upper = FullModel, lower = Base), direction = "both", trace = TRUE)
```

> We will use stepwise forward/backward as it produces a more comprehensive approach to finding the best model

##### Considering possible interactions

```{r}
# considering interactions that would reduce longevity:
# alcohol consumption and AIDS were considered. Alcohol was considered as it is a drug that impairs cognitive function and AIDS was considered due to the fact that it is an STD that will almost always claim its victim's lives

# Full/Reduced Models
# check interactions btwn alcohol and continent
alc_red <- lm(longevity ~ alcohol_trnsfrm + continent, data = na.omit(life_exp_trnsfrm))

alc_int <- lm(longevity ~ alcohol_trnsfrm * continent, data = na.omit(life_exp_trnsfrm))

# anova output of reduced vs interaction model
anova(alc_red, alc_int)

# Full/Reduced Models
# check interactions btwn aids and continent
aids_death_trnsfrm_red <- lm(longevity ~ aids_trnsfrm + continent, data = na.omit(life_exp_trnsfrm))

aids_death_trnsfrm_int <- lm(longevity ~ aids_trnsfrm * continent, data = na.omit(life_exp_trnsfrm))

# anova output of reduced vs interaction model
anova(aids_death_trnsfrm_red, aids_death_trnsfrm_int)

```

> It seems that alcohol may have more of an interaction with continents than does AIDS, which makes sense because some regions have outlawed alcohol whereas AIDS cannot be contained by law

##### Checking the final models to determine if interactions are worth including

```{r}
# checking final models 
m1 <- lm(formula = longevity ~ inc_comp + aids_trnsfrm + tot_exp_trnsfrm + 
         adult_mort + thin_1_19_trnsfrm, data = na.omit(life_exp_trnsfrm))

# output summaries
summary(m1)

# checking for possible interactions
m2 <- lm(formula = longevity ~ inc_comp + aids_trnsfrm + tot_exp_trnsfrm +
           thin_1_19_trnsfrm + adult_mort * alcohol_trnsfrm,
         data = na.omit(life_exp_trnsfrm))

summary(m2)
```

> According to our regression models, inc_comp, aids_trnsfrm, tot_exp_trnsfrm, adult_mort, and thin_1\_19_trnsfrm are the variables that are most asscociated with out outcome of interest. It makes sense that adult mortality will affect life expectancy. It seems that more education (schooling was correlated with income composition) and wealth can afford healthier lifestyles. Seeing how income composition (according to the Human Development Index in terms of income composition of resources - https://www.kaggle.com/datasets/kumarajarshi/life-expectancy-who?resource=download) is so positively correlated with longevity, real world knowledge of the impact of wealth on health is evident. With greater dissemination of technology and information comes the expectation of longer life. In terms of Total Expenditure (which is the government's spending on health), we can see that the more a government spends, the greater the longevity. Mortal diseases such as aids will definitely have a negative impact on longevity as the disease tends to cut life short. Thinness from ages 1-19 will also impact the life expectancy as we can most likely assume that it measures some form of undernourishment. Adult mortality is definitely going to negatively impact longevity as a higher death rate can point to shorter lifespans. The predictor variables explain 86.22% of the variation in Y according to the adjusted $R^2$

### Kfold validation to validate our model

```{r}
# for reproducibility
set.seed(13)

# we will be using kfold validation to validate our model
# getting the length of the dataset
n <- dim(life_exp_trnsfrm)[1]

# for 10 folds
n_fold <- 10

# randomly assign each obs to a fold
folds <- sample(rep(1:n_fold, length = n))

# initialzing vector to hold values
mspe_fold <- rep(0, n_fold)

# run cross validation
for (i in 1:n_fold) {

  train <- na.omit(life_exp_trnsfrm[folds != i, ])
  test <- na.omit(life_exp_trnsfrm[folds == i, ])
  model <- lm(m1, data = train)
  pred <- predict(model, test)
  mspe_fold[i] <- mean((test$longevity - pred)^2)

}

# calculate MSPE
mspe_rmse <- c(avg_MSPE = mean(mspe_fold), avg_RMSE = mean(sqrt(mspe_fold)))

mspe_rmse
```

> Here we see an average MSPE of `r mspe_rmse[1]` and an average RMSE of `r mspe_rmse[2]`

```{r}
# calculate mse
mse_proj <- (2.981)^2

# checking values
mse_proj
```

> From the calculation abomve, we can see that the MSPE and MSE have relatively the same magnitude with a difference of around 1.3

```{r}
# checking the square root of MSE
sqrt(mse_proj)
```

> The RMSEs have the same magnitude and have a difference of about 0.01. These results seem to indicate that the model is valid at least on average

## Model Diagnostics

```{r}
n <- dim(life_exp_trnsfrm)[1]
p <- dim(life_exp_trnsfrm)[2]

# print VIF values to check if any are greater than 10
vif(m1)
```

> There are no results greater than 10

```{r}
# plotting studentized residual plots
ols_plot_resid_stud(m1)
```

> It seems like obervation 111 may be an outlier

```{r}
leverage <- hatvalues(m1)

whichones <- which(leverage>2*p/n)

leverage[whichones]
```

##### Checking influence

```{r}
ols_plot_cooksd_chart(m1)
```

> However the Cook's Distribution tells us differently

```{r}
# sorted since there were observations that were excluded by ols_plot_cooksd_chart and the index values of the chart are not accurate
pf(sort(cooks.distance(m1), decreasing = T)[c(1, 2, 3)], p, n-p)

# to find the threshold for DFFITS
2*sqrt(p/n)

sort(dffits(m1), decreasing = T)[c(1, 2, 3)]

# take a look at the ones that are the same across the tests and then remove (pg 14 of chap 10)

```

> According to the Cook's Distance chart results and the DFFITS results, we may have to remove row 118 instead of row 111.

> Please check Data Cleaning page for columns that were dropped
