---
title: "Project Part 2"
subtitle: "STAT 511"
author: "Moses Sanches, Shuyu Wang, Theodore Choi"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    number_section: false
    highlight: tango
---

```{r}

#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)

```

# Project Part 2

```{r}

#| message: false
# Libraries we used
library(openintro)
library(tidyverse)
library(broom)
library(pastecs)
library(countrycode)

```

## Part A
#### Specify the research question you want to investigate using this data.  
> - Do demographics affect life expectancy?

## Part B
#### 1. Document any initial data cleaning steps you take – e.g. changing variable names, converting characters to factors, etc.

```{r}

# Information regarding data from:
# https://www.kaggle.com/datasets/kumarajarshi/life-expectancy-who?resource=download

# load data (must have file in correct project folder)
life_exp <- read.csv("./data/Life Expectancy Data.csv")
# life_exp

# add continent to dataset
# https://stackoverflow.com/questions/47510141/get-continent-name-from-country-name-in-r
life_exp$continent <- countrycode(sourcevar = life_exp[,"Country"],
                                  origin = "country.name",
                                  destination = "continent")

# Change Variable names for ease of use - including changing to lowercase
life_exp <- life_exp %>% 
  rename(
    country = Country,
    year = Year,
    status = Status,
    longevity = Life.expectancy,
    adult_mort = Adult.Mortality, # number of deaths aged 15 - 60 per 1000 people
    inf_death = infant.deaths, # number  of infant deaths per 1000 people
    alcohol = Alcohol, # Liters of pure alcohol consumed per capita (15 YO +)
    perc_exp = percentage.expenditure, # % Expenditure on health as percentage of GDP per Capita
    hep_b = Hepatitis.B, # % of 1 YO immunized against Hep B
    measles = Measles, # Number of reported cases per 1000 people
    bmi = BMI, # Average BMI for population
    u5_deaths = under.five.deaths, # number of under 5 deaths per 1000
    polio = Polio, # % of 1 YO immunized against polio
    tot_exp = Total.expenditure, # % general gov expenditure on health vs total gov expenditure
    dipth = Diphtheria, # % 1 YO vaccinated TDaP
    aids = HIV.AIDS, # deaths per 1000 live births HIV/AIDS (0-4 years)
    gdp = GDP, # GDP per captia in USD
    pop = Population,
    thin_1_19 = thinness..1.19.years, # % prevalence of thinness among 10 - 19 YO 
    thin_5_9 = thinness.5.9.years, # % prevalence of thinness among 5 - 9 YO 
    inc_comp = Income.composition.of.resources, # Human Development Index - income composition of resources
    school = Schooling # number of years of schooling
  )

# change character columns to factors
life_exp <- as.data.frame(unclass(life_exp), stringsAsFactors = TRUE)

# check classes to see if the changes held
sapply(life_exp, class)

# set the response variable to the Life.expectancy column as y
# this may not be necessary, but was done preemptively for the linear model
y <- life_exp$longevity

# remove the response variable from the dataframe and assign to x
# this may not be necessary, but was done preemptively for the linear model
x <- life_exp[,-4]

```

#### 2. Filter the data to remove any rows that have a missing value for your main response variable. Make note of how many observations this removes. 
```{r}

# check summary to determine if there are nulls
summary(y)

```

> There are 10 null values in the label

```{r}

# remove the null values and assign the new dataframe with removed nulls to new variable
life_exp1 <- life_exp[complete.cases(life_exp$longevity),]

# check for nulls in the summary
summary(life_exp1$longevity)

```

> Removing the null values from the response variable resulted in the removal of 10 rows

#### 3. Narrow down your dataset to a pool of 10 – 20 potential predictor variables. You should document your decisions and justifications. Some things to consider: 
###### a. Which variables are of substantive interest to your research question?

>  ```{r}
life_exp %>%
  group_by(year) %>%
  count(year) %>%
  arrange(desc(n))
```

>  The year(2013), life expectancy, adult mortality, BMI, Alcohol, Hepatitis B, Schooling, Polio, Diphtheria, and Measles. The year 2013 was chosen, since thats where the highest count was. Alcohol, Hepatitis B,  Diphtheria, Measles, Polio were chosen because they are factors that influence health and affect life expectancy. Life expectancy and Adult Mortality help give a big picture of most common ages that both genders are passing away. 

###### b. Which ones aren’t of substantive interest, and why? For example, there may be some administrative variables (e.g. phone number or website) or others deemed irrelevant to the research question. 

```{r}
hist(life_exp$infant.deaths,
     main="Histogram of Infant Deaths",
     xlab="Infant Deaths")
```

```{r}
hist(life_exp$under.five.deaths,
     main="Histogram of Under 5 Deaths",
     xlab="Amount of deaths under 5")
```

```{r}
hist(life_exp$thinness..1.19.years,
     main = "Histogram of Thinness from age 10 to 19",
     xlab = "Age")
```

```{r}
hist(life_exp$thinness.5.9.years,
     main = "Histogram of Thinness from age 5 to 9",
     xlab = "Age")
```

> There were multiple variables that could of been used to help us answer our research question, but there wasn't much variation in the data or it wasn't relevant enough. Some of the variables that couldn't be of use due to the data provided included infant deaths and under 5 deaths due to their values falling under a single range. For example, in under 5 deaths, the data ranged from 0-2500 and the values were only between 0-250. The thinness variables were ones that didnt seem relevant enough to include because they looked vague.


###### c. Are there variables that are unusable due to large amounts of missingness?
```{r}

# show counts of missingness
# first get total row count 
count_rows <- nrow(life_exp)

# Then get sum of NAs
sum_na_gdp <- sum(is.na(life_exp$gdp))
# Then divide by rows to get percentage
per_na_gdp <- (sum_na_gdp / count_rows) * 100

# repeat steps for population and hep b
sum_na_pop <- sum(is.na(life_exp$pop))
per_na_pop <- (sum_na_pop / count_rows) * 100

sum_na_hepb <- sum(is.na(life_exp$hep_b))
per_na_hepb <- (sum_na_hepb / count_rows) * 100

```

> The variables population (missing `r sum_na_pop` rows, which is `r round(per_na_pop, 2)`% missing), hepatitis B(missing `r sum_na_hepb` rows, which is `r round(per_na_hepb, 2)`% missing), and gdp (missing `r sum_na_gdp` rows, which is `r round(per_na_gdp, 2)`% missing) were the most unstable variables due to them having at least 15% of their data missing.

###### d. Are there multiple variables that measure very similar things or should be combined into one variable (e.g. a column for both hours and minutes)? Inspecting a correlation matrix of the data is useful here 

```{r}
cor.life_exp=cor(life_exp[sapply(life_exp1,is.numeric)])
corrplot(cor.life_exp)
```

> Measles could be combined with infant death and deaths under 5 years. Also infant deaths and deaths under 5 years, since they had the highest corrleation.

#### 4. Inspect your missing data again and try to gain understanding about what is missing and why.  
###### a. Read the dataset’s documentation/codebook for insight.

> From the variables we chose to include, hepatitis B, population, and gdp were missing in larger percentages. According to the documentation, lesser known countries provided less data regarding the 3 aformentioned variables.

###### b. Check if there are other values stored as -99 or NA that may indicate missingness. 
```{r}

# check if there are -99s
# NA means there are no -99 values
any(life_exp$longevity == -99)

```

> There are no -99 values in the label

```{r}

# check if there are any 'NA's in the columns
# we used regex to check if there are any characters at all in the column
# NA means there are no characters in the column
sum(str_detect(life_exp$longevity, '[a-zA-Z]')) > 0

```

> There are no character values in the column, which means there are no 'NA' values or combinations of lowercase and uppercase characters that could be portrayed as 'NA' in the column

###### c. Provide a brief summary of the missingness 

> There are no 'NAs' or '99s' in our data, but there blank spaces where there should be values.

##### 5. Document what you decide to do about the missingness, and provide the code for filtering out any observations, if applicable.

> We will leave the blank spaces alone. We won't take out any rows or columns because of the information attached to them.

##### 6. Document any other data cleaning steps you take, if applicable


##### 7. Save your cleaned dataset as a .csv or .RDS (use the function write_csv() or write_rds()). Make sure to name it differently from your original dataset (e.g. data_cleaned or data_for_analysis) and save it in your data folder.  
```{r}

#
#
#
#
#
#

```


## Part C

#### 1. First, provide a histogram of your response variable by itself
```{r}

# plotting histogram of life expectancy 
ggplot(life_exp1, aes(x = longevity)) +
  geom_histogram(binwidth = 5) +
  labs(title = "Histogram of Life Expectancy") +
  ylab("Count of Ages") +
  xlab("Longevity")

```

#### 2. For each continuous predictor

###### a. Create a scatterplot (response on y axis, predictor on x axis)

###### b. Comment on the relationship -- does it appear to be linear? Strong/weak? Positive/negative?

###### c. Are there outliers?

```{r}

# Plotting Alcohol vs Life Expectancy
ggplot(data = life_exp, aes(x = alcohol, y = longevity)) +
  geom_point() +
  labs(title = "Plot of Alcohol vs Life Expectancy") +
  ylab("Longevity") +
  xlab("Alcohol")

```

> It doesn't seem to have any linear relationship, the regression line is probably somewhere near 0. If alcohol and life expectancy are independent, then I believe there are no outliers.

```{r}

# Plotting Percentage Expenditure vs Life Expectancy
ggplot(data = life_exp, aes(x = perc_exp, y = longevity)) +
  geom_point() +
  labs(title = "Plot of Alcohol vs Life Expectancy") +
  ylab("Longevity") +
  xlab("Alcohol")
  
```

> It looks like there is some relationship between percentage expenditure and life expectancy but it's curved instead of linear without outliers.

```{r}
# Plotting BMI vs Life Expectancy
ggplot(data = life_exp, aes(x = bmi, y = longevity)) +
  geom_point() +
  labs(title = "Plot of BMI vs Life Expectancy") +
  ylab("Longevity") +
  xlab("BMI")
```

> BMI and life expectancy has linear relationship and it's positive. There's one obvious outlier.

```{r}
# Plotting Total expenditure vs Life Expectancy
ggplot(data = life_exp, aes(x = tot_exp, y = longevity)) +
  geom_point() +
  labs(title = "Plot of Total expenditure vs Life Expectancy") +
  ylab("Longevity") +
  xlab("Total expenditure")
```

> There's no clear linear relationship between total expenditure and life expectancy.

```{r}
# Plotting HIV.Aids vs Life Expectancy
ggplot(data = life_exp, aes(x = aids, y = longevity)) +
  geom_point() +
  labs(title = "Plot of HIV.Aids vs Life Expectancy") +
  ylab("Longevity") +
  xlab("HIV.Aids")
```

> The relationship between HIV.AIDS and life expectancy is curve as well. No obvious outlier is found.

```{r}
# Plotting GDP vs Life Expectancy
ggplot(data = life_exp, aes(x = gdp,y = longevity)) +
  geom_point() +
  labs(title = "Plot of GDP vs Life Expectancy") +
  ylab("Longevity") +
  xlab("GDP")
```

> The relationship between GDP and life expectancy is curve as well. But there are some outliers.

```{r}
# Plotting Thenniess vs Life Expectancy
ggplot(data = life_exp, aes(x = thin_5_9, y = longevity)) +
  geom_point() +
  labs(title = "Plot of Thenniess vs Life Expectancy") +
  ylab("Longevity") +
  xlab("Thenniess")
```

> It seems that thenniess5.9.years and life expectancy has some relationship but not clear, but it should be curve but not linear.

```{r}
# Plotting Income.composition.of.resources vs Life Expectancy
ggplot(data = life_exp, aes(x = inc_comp, y = longevity)) +
  geom_point() +
  labs(title = "Plot of Income_comp vs Life Expectancy") +
  ylab("Longevity") +
  xlab("Income_comp")
```

> There is positive linear relationship between income.composition.of.resources and life expectancy. The life line are the outliers.

```{r}
# Plotting Schooling vs Life Expectancy
ggplot(data = life_exp, aes(x = school, y = longevity)) +
  geom_point() +
  labs(title = "Plot of Schooling vs Life Expectancy") +
  ylab("Longevity") +
  xlab("Schooling")
```

> There is positive linear relationship between Schooling and life expectancy as well. The life line are still the outliers.

#### 3. For each categorical predictor

There are two categorical predictor. Country and Status. 2938 observations in each predictor.

```{r}

# plot boxplot of life expectancy across continents
ggplot(life_exp,
       aes(x = continent,
           y = longevity,
           fill = continent)) +
  geom_boxplot(outlier.color = "black",
               outlier.shape = 16,
               outlier.size = 2,
               notch = FALSE) +
  labs(title = "Boxplot of Life Expectancy Across Continents") +
  ylab("Longevity") +
  xlab("Continents")

```

> There appear to be differences in Y across the different levels of X, and there are some plots which are outliers.

```{r}
ggplot(life_exp,
       aes(x = status,
           y = longevity,
           fill = status)) +
  geom_boxplot(outlier.color = "black",
               outlier.shape = 16,
               outlier.size = 2,
               notch = FALSE) +
  labs(title = "Boxplot of Life Expectancy Across Status") +
  ylab("Longevity") +
  xlab("Status")
```

> There appear to be differences in Y across the different levels of X, and there are some outliers in developing countries.
