---
title: "Project Appendix
"
subtitle: "STAT 511"
author: "Moses Sanches, Shuyu Wang, Theodore Choi"
date: "`r Sys.Date()`"
format: 
  html:
    self-contained: true
    toc: true
    toc_float: true
    number-section: false
    highlight: tango
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE)

```

```{r}
#| message: false
#| warning: false
# Libraries we used
library(openintro)
library(tidyverse)
library(dplyr)
library(scales)
library(leaps)
library(MASS)
library(olsrr)
library(car)
```

### Preliminary analysis using basic visualizations
```{r}
# getting summaries of the status counts
# string values did not return summary statistics
summary_num <- stat.desc(select_if(life_exp, is.numeric))
round(summary_num, 2)

# frequency of status by country
table(life_exp$Continent, life_exp$Status) 
```

#### B. Produce visual summaries (one for each variable based on variable type) – boxplots/histograms or bar charts.
```{r}
# Getting counts of the Status (Developing vs developed)
counts_status <- table(life_exp$Status)

# Plotting a bar chart of the counts
barplot(counts_status,
        main = "Status Counts",
        ylab = "Counts",
        xlab = "Status")

# Getting frequency of Adult Mortality
# Defined as:
# "Adult Mortality Rates of both sexes
# (probability of dying between 15 and 60 years per 1000 population)"
hist(life_exp$Adult.Mortality,
     main = "Histogram of Adult Mortality",
     xlab = "Adult Mortality")

# Consumption of Alcohol per capita (15+) in liters
hist(life_exp$Alcohol,
     main = "Histogram of Alcohol",
     xlab = "Alcohol Consumption")

# Hepatitis B (HepB) immunization coverage among 1-year-olds (%)
hist(life_exp$Hepatitis.B,
     main = "Histogram of Hepatits B",
     xlab = "Infected with Hepatitis B")

# Year
hist(life_exp$Year,
     main = "Histogram of year",
     xlab = "Years")

# Plotting a bar chart of the  Continent counts
# (too many countries to plot cleanly)
counts_continent <- table(life_exp$Continent)
barplot(counts_continent,
        main = "Count of Continents",
        ylab = "Counts",
        xlab = "Continents")

# infant deaths
hist(life_exp$infant.deaths,
     main = "Histogram of Infant Deaths",
     xlab = "Infant Deaths")

# percentage expenditure 
hist(life_exp$percentage.expenditure,
     main = "Histogram of Percentage Expenditure",
     xlab = "Percent Expenditure")

# Measles
hist(life_exp$Measles,
     main = "Histogram of Measles",
     xlab = "Infected with Measles")

# BMI
hist(life_exp$BMI,
     main = "Histogram of BMI",
     xlab = "BMI")

# under 5 deaths
hist(life_exp$under.five.deaths,
     main = "Histogram of Under 5 Deaths",
     xlab = "Amount of deaths under 5")

# Polio
hist(life_exp$Polio,
     main = "Histogram of Polio Occurence",
     xlab = "Polio Occurrence")

# Total Expenditure
hist(life_exp$Total.expenditure,
     main = "Histogram of Total Expenditure",
     xlab = "Total Expenditure")

# Diphtheria
hist(life_exp$Diphtheria,
     main = "Histogram of Frequency of Diphtheria",
     xlab = "Diphtheria")

# AIDS
hist(life_exp$HIV.AIDS,
     main = "Histogram of Frequency of HIV AIDS",
     xlab = "HIV AIDS")

# GDP
hist(life_exp$GDP,
     main = "Histogram of GDP",
     xlab = "GDP")

# Population
hist(life_exp$Population,
     main = "Histogram of Population",
     xlab = "Population")

# Thinness(1-19 years):
# Prevalence of thinness among children and adolescents for Age 10 to 19 (%)
hist(life_exp$thinness..1.19.years,
     main = "Histogram of Thinness from age 10 to 19",
     xlab = "Age")

# Thinness(5-9 years):Prevalence of thinness among children for Age 5 to 9(%)
hist(life_exp$thinness.5.9.years,
     main = "Histogram of Thinness from age 5 to 9",
     xlab = "Age")

# Income Composition of Resources:
# Human Development Index in terms of income composition of resources
# (index ranging from 0 to 1)
hist(life_exp$Income.composition.of.resources,
     main = "Histogram of Resource Composition",
     xlab = "Index")

# Schooling:Number of years of Schooling(years)
hist(life_exp$thinness.5.9.years,
     main = "Histogram of Years of Schooling",
     xlab = "Years")

# plotting a histogram of the response variable (life expectancy)
hist(y,
     main = "Histogram of Life Expectancy",
     xlab = "Age")

# boxplot of life expectancy across the continents
life_exp %>%
  ggplot(aes(x = Continent,
             y = Life.expectancy,
             fill = Continent)) +
  geom_boxplot(outlier.color = "black",
               outlier.shape = 16,
               outlier.size = 2,
               notch = FALSE)
```

#### C. Are there any potential issues with any of the data?
> The data still has some missing values and can be a little difficult to interpret - use of the data descriptors found in the link is highly recommended. The Life Expectancy column (response variable) is skewed left, so some transformations may have to be performed to ensure a normal curve. We can also see skewed distributions across the variables (Histogram of Infant Deaths and Histogram of Under 5 Deaths among others), possible missing data (as seen in Histogram of Total Expenditure), as well as possible outliers (as seen in Histogram of Population,the Histograms of thinness). Another interesting issue that we can see is a bimodal distribution (as seen in the Histogram of BMI). There may also be issues with the frequency of the number of years for the early 2000s as there is a lot of data for 2000.

```{r}

# import data
filepath_2013 <- "./data/longevity Data Cleaned Version 2013 TFM.rds"

life_exp_pt4 <- readRDS(filepath_2013)

```


```{r}

n <- dim(life_exp_pt4)[1]
p <- dim(life_exp_pt4)[2]


m1 <- lm(formula = life_exp_2013_response ~ inc_comp + aids_trnsfrm +
    tot_exp_trnsfrm + adult_mort + perc_exp_trnsfrm + gdp_trnsfrm,
    data = life_exp_pt4)

# print VIF values to check if any are greater than 10
vif(m1)

ols_plot_resid_stud(m1)

leverage <- hatvalues(m1)

whichones <- which(leverage>2*p/n)

leverage[whichones]

ols_plot_cooksd_chart(m1)

# sorted since there were observations that were excluded by ols_plot_cooksd_chart and the index values of the chart are not accurate
pf(sort(cooks.distance(m1), decreasing = T)[c(1, 2, 3)], p, n-p)

# to find the threshold for DFFITS
2*sqrt(p/n)

sort(dffits(m1), decreasing = T)[c(1, 2, 3)]

# take a look at the ones that are the same across the tests and then remove (pg 14 of chap 10)

```













